---
title: "xEnrich: from gene list to pathway, organ, and cell type"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{xEnrich: from gene list to pathway, organ, and cell type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
library(xEnrich)
```

## The problem

Standard enrichment analysis returns dozens of nominally significant
pathways, but doesn't distinguish independent biological signals from
redundant re-descriptions of the same gene set.  It also can't tell you
*where* in the body a pathway is active.

xEnrich addresses both gaps:

- **Track A** identifies *which* pathways are altered and separates
  independent signals from redundant shadows using conditional mutual
  information.
- **Track B** identifies *where* in the body the gene list is most
  relevant and drills down to cell types.

This vignette demonstrates the full workflow on synthetic data.

## Setup: synthetic data with known ground truth

```{r synthetic-data}
set.seed(42)
n_samples <- 120
n_genes   <- 500
genes     <- paste0("G", seq_len(n_genes))

# Two independent latent factors drive two clusters of genes
factor_A <- rnorm(n_samples)
factor_B <- rnorm(n_samples)

expr <- matrix(
  rnorm(n_samples * n_genes),
  nrow     = n_samples,
  dimnames = list(paste0("S", seq_len(n_samples)), genes)
)

# Inject signal into two gene blocks
expr[, 1:40]    <- expr[, 1:40]    + 2.5 * factor_A
expr[, 300:340] <- expr[, 300:340] + 2.0 * factor_B

# Phenotype: weighted sum of both factors + noise
phenotype <- 0.7 * factor_A + 0.5 * factor_B + rnorm(n_samples, sd = 0.4)

# Pathway definitions: two true signals, one shadow, one noise
gene_sets <- list(
  immune_response   = genes[1:40],           # true signal (factor A)
  immune_regulation = genes[1:35],           # redundant shadow of immune_response
  solute_transport  = genes[300:340],        # true signal (factor B)
  cell_adhesion     = genes[200:230],        # noise
  lipid_metabolism  = genes[400:430]         # noise
)
```

## Track A: pathway-phenotype association

### Step 1: run_info_assoc()

Mutual information between each pathway's mean activity and the
phenotype, calibrated by permutation:

```{r assoc}
assoc <- run_info_assoc(
  expr      = expr,
  phenotype = phenotype,
  gene_sets = gene_sets,
  n_perm    = 500,
  seed      = 42
)

assoc[, c("set", "set_size", "MI_bits", "z_score", "p_value", "padj")]
```

Three pathways are significant: `immune_response`, `immune_regulation`
(its shadow), and `solute_transport`. The noise pathways are correctly
non-significant.

### Step 2: run_assoc_redundancy_selection()

Which of these are independent findings?

```{r redundancy}
sel <- run_assoc_redundancy_selection(
  assoc_results = assoc,
  expr          = expr,
  phenotype     = phenotype,
  gene_sets     = gene_sets,
  alpha         = 0.05,
  seed          = 42
)

sel[, c("step", "pathway", "marginal_mi", "conditional_mi",
        "redundancy_ratio", "cumulative_bits")]
```

The redundancy ratio reveals the answer:

- `immune_response` and `solute_transport` have ratios near 1 — they
  are independent signals.
- `immune_regulation` has a ratio near 0 — its association with the
  phenotype is entirely explained by `immune_response`.

### Step 3: classify_assoc_selection()

Translate ratios into plain-language actions:

```{r classify}
classified <- classify_assoc_selection(sel)

classified[, c("pathway", "marginal_strength", "signal_class", "action")]
```

### Step 4: compute_pathway_activity()

Extract per-sample activity scores for the selected pathways, ready for
heatmaps, mediation models, or external regression:

```{r activity}
act <- compute_pathway_activity(
  expr      = expr,
  gene_sets = gene_sets[sel$pathway],
  score     = "mean_z"
)

cat("Activity matrix:", nrow(act), "samples x", ncol(act), "pathways\n")
cat("Correlation with phenotype:\n")
round(cor(act, phenotype), 3)
```

### Visualise: information gain per selection step

```{r gains-plot}
plot_gains(sel)
```

## Track A: gene-list enrichment

When you have only a gene list (no expression matrix), use the
gene-list lane.

### Enrichment + redundancy in gene space

```{r gene-list-lane}
# Simulate a DE gene list
gene_list <- unique(c(
  sample(genes[1:40],    30),   # enriched in immune_response
  sample(genes[300:340], 20),   # enriched in solute_transport
  sample(genes[100:200], 5)     # a few noise genes
))

# MI-scored enrichment
enrich <- run_info_enrichment(
  gene_list = gene_list,
  gene_sets = gene_sets,
  universe  = genes,
  n_perm    = 500,
  seed      = 42
)

enrich[, c("set", "overlap", "set_size", "info_bits", "ratio",
           "padj_emp", "padj_enrich")]
```

Note how `immune_regulation` is significant by hypergeometric p-value
(`padj_enrich`) but has lower MI (`info_bits`) than `immune_response`
because its genes are a subset — less informative.

```{r gene-space-redundancy}
# Redundancy selection in gene space
sel_gl <- run_redundancy_selection(
  gene_list      = gene_list,
  gene_sets      = gene_sets,
  universe       = genes,
  initial_scores = setNames(enrich$info_bits, enrich$set),
  seed           = 42
)

sel_gl[, c("step", "pathway", "marginal_mi", "conditional_mi",
           "redundancy_ratio")]
```

Gene-space CMI correctly identifies `immune_regulation` as redundant
with `immune_response`.

## Track A: GSEA lane

When you have a full ranked list:

```{r gsea-lane}
# Gene-level statistics: correlation with phenotype
gene_stats <- setNames(
  apply(expr, 2, function(g) cor(g, phenotype)),
  genes
)

batch <- run_batch_gsea(
  gene_sets  = gene_sets,
  gene_stats = gene_stats,
  n_perm     = 500,
  adaptive   = TRUE,
  seed       = 42
)

batch$summary[, c("pathway", "enrichment_score", "p_value", "padj",
                   "leading_edge_size")]
```

```{r gsea-redundancy, eval = nrow(batch$summary[!is.na(batch$summary$padj) & batch$summary$padj < 0.05, ]) >= 2}
# Redundancy on leading edges
sel_gsea <- run_gsea_redundancy_selection(
  gsea_results = batch$results,
  gene_stats   = gene_stats,
  alpha        = 0.05,
  seed         = 42
)

# Classify
classified_gsea <- classify_gsea_selection(sel_gsea)
classified_gsea[, c("pathway", "signal_class", "action")]
```

## Adaptive permutation p-values

xEnrich reports confidence intervals on permutation p-values. Here is
one pathway in detail:

```{r adaptive-detail}
res <- run_enrichment(
  gene_list     = gene_sets$immune_response,
  gene_stats    = gene_stats,
  method        = "gsea",
  n_perm        = 500,
  adaptive      = TRUE,
  adaptive_mode = "decision",
  alpha         = 0.05,
  seed          = 42,
  keep_trace    = TRUE
)

inf <- res$inference$res
cat(sprintf(
  "ES = %.3f\np = %.4f\nCI = [%.4f, %.4f]\nPermutations used: %d\nDecision: %s\n",
  res$enrichment_score, inf$p,
  inf$p_ci_low, inf$p_ci_high,
  inf$B, inf$decision_alpha
))
```

The Wilson confidence interval makes the uncertainty in the p-value
estimate explicit. Standard tools report only the point estimate.

## Track B: location enrichment

Track B requires Human Protein Atlas and Tabula Sapiens reference
data. These are downloaded once via `load_reference()` and cached
locally:

```r
# Not run — requires network access on first call
hpa_rna       <- load_reference("rna_hpa", source = "hpa")[[1]]
tabula_kidney <- load_reference("kidney",  source = "tabula")[[1]]
```

### Tissue-level enrichment

```r
# Tissue enrichment using HPA RNA data
tissue_result <- enrich_by_hpa(
  gene_list      = my_gene_list,
  reference_df   = hpa_rna,
  group_col      = "Tissue",
  expression_col = "nTPM",
  method         = "hypergeometric",
  cutoff_type    = "quantile",
  cutoff_value   = 0.9,
  universe       = my_universe,
  universe_type  = "provided"
)
```

### Full pipeline: organ → cell type

```r
# All-in-one: HPA tissue enrichment -> Tabula cell-type drill-down
pipeline <- run_location_pipeline(
  gene_list   = my_gene_list,
  hpa_data    = hpa_rna,
  tabula_data = list(kidney = tabula_kidney),
  hpa_type    = "rna",
  seed        = 42
)

print(pipeline)    # organ-level summary
summary(pipeline)  # flat data frame of top cell types
```

## Convergence: WHAT + WHERE

The gene list bridges both tracks:

```
gene list → Track A: immune_response     → WHAT  (pathway)
         → Track B: cerebral cortex     → WHERE (organ)
                    → neurons           → WHERE (cell type)
         ↓
         Together: "immune response / cerebral cortex / neurons"
```

This convergence — naming the biology, the organ, and the cell type
from a single gene list — is what xEnrich enables that no single
existing tool provides.

## Session info

```{r session}
sessionInfo()
```
